Manuscript Figures for Genome Biology

### 2018-09-11 manuscript_figures/manuscript_figures.Rmd
Add alternate color scales for Figures S4 and S9 - log-scale % and log2 Ratio

# calculate percentage diff as percentage of the # of rejections found by 
# using no covariate. If no rejections are found using no covariate, and
# rejections are found with the covariate, set to 100%. 
dftab <- left_join(inftab, uninftab, by = "id") %>%
  dplyr::mutate(diff_inf = nrejects - uninf) %>%
  dplyr::mutate(diff_inf_percent = diff_inf / uninf) %>%
  dplyr::mutate(diff_inf_percent = ifelse(uninf==0 & nrejects > 0, 1, 
                                          diff_inf_percent)) %>%
  dplyr::mutate(method = factor(method, levels = rowOrder)) %>%
  dplyr::mutate(casestudy = factor(casestudy, levels = colOrder)) %>%
  dplyr::group_by(casestudy, cov) %>%
  complete(method, dataset)

maxVals <- dftab %>% 
  group_by(dataset,cov) %>%
  summarize(top = max(diff_inf_percent, na.rm=TRUE))

dftab <- left_join(dftab, maxVals, by = c("dataset", "cov")) %>%
  mutate(top = ifelse(diff_inf_percent == top, diff_inf_percent, NA))

# if no change, set diff_inf_percent to zero, even if no rejections in either case
dftab <- dftab %>% 
  mutate(diff_inf_percent = ifelse(nrejects == uninf, 0, diff_inf_percent))

trim_tails <- function(range = c(-Inf, Inf)) scales::trans_new("trim_tails", 
                transform = function(x) {
                  force(range)
                  desired_breaks <- scales::extended_breaks(n = 7)(x[x >= range[1] & x <= range[2]])
                  break_increment <- diff(desired_breaks)[1]
                  x[x < range[1]] <- range[1] - break_increment
                  x[x > range[2]] <- range[2] + break_increment
                  x
                },
                inverse = function(x) x,

                breaks = function(x) {
                  force(range)
                  scales::extended_breaks(n = 7)(x)
                },
                format = function(x) {
                  force(range)
                  x[1] <- paste("<", range[1])
                  x[length(x)] <- paste(">", range[2])
                  paste0(x,"%")
                })

Fig_impactofinfcov <- ggplot(dftab, aes(x = dataset, y = method, fill = diff_inf_percent*100)) + 
        geom_tile() + 
        geom_text(aes(label = ifelse(is.na(top), "", #adaptive text lab color
                                     ifelse(top > 0.7, scales::percent(top), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(top), "", 
                                     ifelse(top <= 0.7, scales::percent(top), ""))), 
                  color = "black", size = 2)+
        scale_fill_gradientn("% Change in\nrejections",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-20, 100),
                             values = scales::rescale(c(-20, 0, 100)), 
                             trans = trim_tails(range = c(-20, 100)),
                             oob = scales::squish) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        facet_grid( ~ casestudy + cov, scales = "free", space = "free") +
        ggtitle("Impact of informative covariate on number of rejections") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.99, 0.95), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=5)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-8, l=10))

@@

@@

```{r}
dftab <- dftab %>%
  group_by(casestudy, method) %>%
  summarize(ndatasets = sum(!is.na(diff_inf_percent)),
            diff_inf_percent = mean(diff_inf_percent, na.rm = TRUE))

maxVals <- dftab %>% 
  group_by(casestudy) %>%
  summarize(top = max(diff_inf_percent, na.rm=TRUE),
            maxd = max(ndatasets))

dftab <- left_join(dftab, maxVals, by = c("casestudy")) %>%
  mutate(top = ifelse(diff_inf_percent == top, diff_inf_percent, NA)) %>%
  ungroup() %>%
  mutate(casestudy = factor(casestudy, levels = colOrder)) %>%
  mutate(casestudyN = paste0(casestudy, "(", maxd, ")")) %>%
  mutate(casestudyN = factor(casestudyN, 
                             levels = unique(casestudyN)[unique(as.numeric(casestudy))]))

Fig_impactofinfcov <- ggplot(dftab, aes(x = casestudy, y = method, fill = diff_inf_percent*100)) + 
        geom_tile() + 
        scale_fill_gradientn("% Change in\nrejections",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-20, 100),
                             values = scales::rescale(c(-20, 0, 100)), 
                             trans = trim_tails(range = c(-20, 100)),
                             oob = scales::squish) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        ggtitle("Impact of informative covariate") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1)) +
        geom_text(aes(label = ifelse(is.na(top), "", #adaptive text lab color
                                     ifelse(top > 0.7, scales::percent(top), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(top), "", 
                                     ifelse(top <= 0.7, scales::percent(top), ""))), 
                  color = "black", size = 2)  +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.945, 0.94), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=5)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-8, l=10))

@@

>>>>

# calculate percentage diff as percentage of the # of rejections found by 
# using no covariate. If no rejections are found using no covariate, and
# rejections are found with the covariate, set to 100%. 
dftab <- left_join(inftab, uninftab, by = "id") %>%
  dplyr::mutate(diff_inf = nrejects - uninf) %>%
  dplyr::mutate(diff_inf_percent = diff_inf / uninf) %>%
  dplyr::mutate(diff_inf_ratio = log2(nrejects / uninf)) %>%
  dplyr::mutate(diff_inf_percent = ifelse(uninf==0 & nrejects > 0, 1, 
                                          diff_inf_percent)) %>%
  dplyr::mutate(diff_inf_ratio = ifelse(uninf==0 & nrejects > 0, 1, 
                                        diff_inf_ratio)) %>%
  dplyr::mutate(diff_inf_ratio = ifelse(uninf==0 & nrejects==0, 0, 
                                        diff_inf_ratio)) %>%
  dplyr::mutate(method = factor(method, levels = rowOrder)) %>%
  dplyr::mutate(casestudy = factor(casestudy, levels = colOrder)) %>%
  dplyr::group_by(casestudy, cov) %>%
  complete(method, dataset)

maxVals <- dftab %>% 
  group_by(dataset,cov) %>%
  summarize(top = max(diff_inf_percent, na.rm=TRUE),
            topR = max(diff_inf_ratio, na.rm =TRUE))

dftab <- left_join(dftab, maxVals, by = c("dataset", "cov")) %>%
  dplyr::mutate(top = ifelse(diff_inf_percent == top, diff_inf_percent, NA),
         topR = ifelse(diff_inf_ratio == topR, diff_inf_ratio, NA))

# if no change, set diff_inf_percent to zero, even if no rejections in either case
dftab <- dftab %>% 
  dplyr::mutate(diff_inf_percent = ifelse(nrejects == uninf, 0, diff_inf_percent))

# custom symmetric log scale for percent
log_neg_percent <- function()
              scales::trans_new("log_neg_percent", 
                transform = function(x) {
                  x <- ifelse(x > 0, log(x + 1), -log(-x + 1))
                  x
                },
                inverse = function(x) {
                  x <- ifelse(x > 0, exp(x) - 1, -exp(-x) + 1)
                  x
                },
                breaks = function(x) {
                  if (max(x) > 200){
                    c(-20, -5, 0, 5, 25, 100, 300)
                  }else if (max(x) > 100){
                    c(-20, -5, 0, 5, 25, 100, 200)
                  }else{
                    c(-20, -5, 0, 5, 25, 100)
                  }
                },
                format = function(x) {
                  paste0(x,"%")
                })

# log-scale color percent fill
topRange <- max(dftab$diff_inf_percent, na.rm = TRUE)*100
botRange <- min(dftab$diff_inf_percent, na.rm = TRUE)*100

Fig_impactofinfcov <- ggplot(dftab, aes(x = dataset, y = method, fill = diff_inf_percent*100)) + 
        geom_tile() + 
        geom_text(aes(label = ifelse(is.na(top), "", #adaptive text lab color
                                     ifelse(top > 0.7, scales::percent(top), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(top), "", 
                                     ifelse(top <= 0.7, scales::percent(top), ""))), 
                  color = "black", size = 2)+
        scale_fill_gradientn("% Change in\nrejections",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-20, topRange),
                             values = scales::rescale(c(-transform(20), 0, transform(topRange))), 
                             trans = log_neg_percent()) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        facet_grid( ~ casestudy + cov, scales = "free", space = "free") +
        ggtitle("Impact of informative covariate on number of rejections") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.99, 0.95), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=2)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-6, l=10))

@@

# log-scale color log2 ratio fill
topRange <- max(dftab$diff_inf_ratio, na.rm = TRUE)
botRange <- min(dftab$diff_inf_ratio, na.rm = TRUE)

Fig_impactofinfcov <- ggplot(dftab, aes(x = dataset, y = method, fill = diff_inf_ratio)) + 
        geom_tile() + 
        geom_text(aes(label = ifelse(is.na(topR), "", #adaptive text lab color
                                     ifelse(topR > 1, signif(topR,2), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(topR), "", 
                                     ifelse(topR <= 1, signif(topR,2), ""))), 
                  color = "black", size = 2)+
        scale_fill_gradientn("Fold Change\n(log2)",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-0.5, topRange), 
                             values = scales::rescale(c(-0.5, 0, topRange))) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        facet_grid( ~ casestudy + cov, scales = "free", space = "free") +
        ggtitle("Impact of informative covariate on number of rejections") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.99, 0.95), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=0)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-2, l=10))

Fig_impactofinfcov
ggsave(file.path(outdir, "Figure_impactofinfcov_fc.pdf"), width = 15.5, height = 4.25)

@@

```{r}
dftab <- dftab %>%
  group_by(casestudy, method) %>%
  summarize(ndatasets = sum(!is.na(diff_inf_percent)),
            diff_inf_percent = mean(diff_inf_percent, na.rm = TRUE), 
            diff_inf_ratio = mean(diff_inf_ratio, na.rm = TRUE))

maxVals <- dftab %>% 
  group_by(casestudy) %>%
  summarize(top = max(diff_inf_percent, na.rm=TRUE),
            topR = max(diff_inf_ratio, na.rm=TRUE),
            maxd = max(ndatasets))

dftab <- left_join(dftab, maxVals, by = c("casestudy")) %>%
  mutate(top = ifelse(diff_inf_percent == top, diff_inf_percent, NA)) %>%
  mutate(topR = ifelse(diff_inf_ratio == topR, diff_inf_ratio, NA)) %>%
  ungroup() %>%
  mutate(casestudy = factor(casestudy, levels = colOrder)) %>%
  mutate(casestudyN = paste0(casestudy, "(", maxd, ")")) %>%
  mutate(casestudyN = factor(casestudyN, 
                             levels = unique(casestudyN)[unique(as.numeric(casestudy))]))

# log-scale color percent fill
topRange <- max(dftab$diff_inf_percent, na.rm = TRUE)*100
botRange <- min(dftab$diff_inf_percent, na.rm = TRUE)*100

Fig_impactofinfcov <- ggplot(dftab, aes(x = casestudy, y = method, fill = diff_inf_percent*100)) + 
        geom_tile() + 
        scale_fill_gradientn("% Change in\nrejections",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-20, topRange),
                             values = scales::rescale(c(-transform(20), 0, transform(topRange))), 
                             trans = log_neg_percent()) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        ggtitle("Impact of informative covariate") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1)) +
        geom_text(aes(label = ifelse(is.na(top), "", #adaptive text lab color
                                     ifelse(top > 0.7, scales::percent(top), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(top), "", 
                                     ifelse(top <= 0.7, scales::percent(top), ""))), 
                  color = "black", size = 2)  +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.945, 0.94), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=2)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-6, l=10))

@@

# log-scale color log2 ratio fill
topRange <- max(dftab$diff_inf_ratio, na.rm = TRUE)
botRange <- min(dftab$diff_inf_ratio, na.rm = TRUE)

Fig_impactofinfcov <- ggplot(dftab, aes(x = casestudy, y = method, fill = diff_inf_ratio)) + 
        geom_tile() + 
        scale_fill_gradientn("Fold Change\n(log2)",
                             colors = c("darkblue","white","darkred"), 
                             limits = c(-0.5, topRange), 
                             values = scales::rescale(c(-0.5, 0, topRange))) +
        theme_bw() +
        xlab("Case Study") +
        ylab("Method") +
        ggtitle("Impact of informative covariate") +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1)) +
        geom_text(aes(label = ifelse(is.na(topR), "", #adaptive text lab color
                                     ifelse(topR > 1, signif(topR, 2), ""))), 
                  color = "white", size = 2) +
        geom_text(aes(label = ifelse(is.na(topR), "", 
                                     ifelse(topR <= 1, signif(topR, 2), ""))), 
                  color = "black", size = 2)  +
        theme(axis.text.x = element_text(angle = 90, vjust = 1/2, hjust = 1),
              legend.position = c(0.945, 0.94), legend.justification = c(1, 1),
              legend.background = element_rect(fill = scales::alpha("white", 3/4), color = 'black'),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7, angle=45, margin = margin(t=2)),
              legend.title = element_text(size = 9, margin=margin(t=-10)),
              legend.margin = margin(t=2, r=10, b=-6, l=10))

Fig_impactofinfcov
ggsave(file.path(outdir, "Figure_impactofinfcov_avg_fc.pdf"), width = 5, height = 4)
